Meta analysis of differential expression results from AMP-AD (Control vs AD)

### 2016-08-17 code/Rmd/MetaAnalysis.Rmd
Minor modifications

@@

@@

### Get covariates/sample metadata from synapse
```{r covariates}
covariates.id = c('ROSMAP' = 'syn6114444',
                  'MSBB' = 'syn6131421',
                  'MAYO' = 'syn6128267')

@@

covariates = rbindlist(covariates, use.names = T, fill = T, idcol = 'Study')

```

### Get logCPM counts from synapse
```{r counts}
logcpm.id = c('ROSMAP' = 'syn6114447' ,
              'MSBB' = 'syn6131419',
              'MAYO' = 'syn6128269')

@@

### Download differential expression data
```{r download.data}
diffexp.id = c('ROSMAP' = 'syn6114453',
               'MSBB' = 'syn6131423',
               'MAYO' = 'syn6128273')

@@

diffexp$ROSMAP = diffexp$ROSMAP %>%
  filter(Comparison == 'cogdx1-cogdx4') %>%
  dplyr::select(Gene.ID, logFC, P.Value, adj.P.Val) %>%
  dplyr::mutate(Tissue = 'DLPFC', Contrast = 'NCI-AD')

diffexp$MSBB = diffexp$MSBB %>%
  dplyr::filter(Name == 'ND-SD') %>%
  dplyr::select(Gene.ID, logFC, P.Value, adj.P.Val, BrainRegion, Name) %>%
  dplyr::mutate(BrainRegion = factor(BrainRegion, labels = c('BM_10' = 'FP', 'BM_22' = 'STG', 
                                                             'BM_36' = 'PHG', 'BM_44' = 'IFG'))) %>%
  dplyr::rename(Tissue = BrainRegion, Contrast = Name)

diffexp$MAYO = diffexp$MAYO %>%
  dplyr::filter(Name == 'Control-AD') %>%
  dplyr::select(genes, logFC, P.Value, adj.P.Val, BrainRegion, Name) %>%
  dplyr::rename(Tissue = BrainRegion, Gene.ID = genes, Contrast = Name)

diffexp = rbindlist(diffexp, use.names = T, fill = T, idcol = 'Study') %>%
  tidyr::separate(Gene.ID, c('ensembl_gene_id','position'),sep = '\\.')

@@

# Add hgnc_symbols to the diffexp matrix
diffexp = diffexp %>%
  left_join(Ensemble2HGNC)
```

@@

### Meta Analysis
```{r metaanalysis}
P.Value = diffexp %>% 
  # filter(Tissue %in% c('DLPFC', 'PHG', 'TCX')) %>%
  dplyr::select(ensembl_gene_id, P.Value, Tissue, Contrast) %>%
  tidyr::unite(Tissue.Contrast, Tissue, Contrast, sep = '_') %>%
  tidyr::spread(Tissue.Contrast, P.Value)
P.Value[is.na(P.Value)] = 1

@@

metaAnal = join_all(list(min.pval, max.pval, fb.pval), type = 'full')
tmp = metaAnal %>% gather(method, pval, -ensembl_gene_id) %>% group_by(method) %>% filter(pval <= 0.05, method == 'fb.pval') %>%
  left_join(Ensemble2HGNC) %>%
  dplyr::select(hgnc_symbol) %>%
  unique
```

```{r plotDE}
tmp = lapply(DEXP, function(x){
  x$observed.P.val = p.adjust(x$P.Value, 'fdr')
  x = arrange(x, observed.P.val)
  x = mutate(x, expected.P.val = seq(0,1,1/(dim(x)[1]-1)))
  x$dexp = 0;
  x$dexp[x$observed.P.val <= 0.05 & x$logFC > 0] = 1;
  x$dexp[x$observed.P.val <= 0.05 & x$logFC < 0] = -1;
  dplyr::select(x, observed.P.val, expected.P.val, dexp)
})

p = list()
for (i in 1:length(tmp)){
  p[[i]] = ggplot(tmp[[i]], aes(x = -log10(expected.P.val), y = -log10(observed.P.val))) + geom_point(aes(color = factor(dexp))) 
  p[[i]] = p[[i]] + scale_colour_manual(values=c('green','grey','red')) + theme(legend.position="none")
  x = -log10(tmp[[i]]$expected.P.val)
  x[4]-x[1]/
  p[[i]] = p[[i]] + geom_abline(slope=1, intercept=0) + ggtitle(names(tmp)[i])
}
multiplot(plotlist = p, cols = 4)
```

```{r uniqGenes}
GNames = DEXP[[1]]$hgnc_symbol
for (i in 2: length(DEXP)){
  GNames = intersect(GNames, DEXP[[i]]$hgnc_symbol)
}
GNames = GNames[!is.na(GNames)]
GNames = GNames[GNames != '']

DEXP = lapply(DEXP, function(dexp, GNames){
  dexp = dexp %>%
    dplyr::filter(hgnc_symbol %in% GNames) %>%
    group_by(hgnc_symbol) %>%
    dplyr::summarise(P.Value = min(P.Value)) %>%
    data.frame
  rownames(dexp) = dexp$hgnc_symbol
  dexp = dexp[GNames,]
  }, GNames)

DEXP = mapply(function(dexp, dataName){
  colnames(dexp) = c('hgnc_symbol', dataName)
  return(dexp)
}, DEXP, names(DEXP), SIMPLIFY = F)

DEXP = join_all(DEXP, by = "hgnc_symbol")
```
Overall `r dim(DEXP)[1]` genes in `r dim(DEXP)[2]-1` studies are analysed
### Differential expression analysis
```{r diffExp, results = 'asis'}
Adj.P.Value = apply(DEXP[,-(1)], 2, p.adjust, method = "fdr")
rownames(Adj.P.Value) = DEXP$hgnc_symbol

writeLines('Total number of genes that are differentially expressed at a FDR <= 0.05 are')
tmp = (data.frame(colSums(Adj.P.Value <= 0.05)))
colnames(tmp) = 'No. of Genes'
kable(tmp)

writeLines('Genes that are differentially expressed at a FDR <= 0.05 in 8/8 datasets')
paste(rownames(Adj.P.Value)[which(rowSums(Adj.P.Value<=0.05)==8)], collapse = ',')

writeLines('Genes that are differentially expressed at a FDR <= 0.05 in 7/8 datasets')
paste(rownames(Adj.P.Value)[which(rowSums(Adj.P.Value<=0.05)==7)], collapse = ',')

writeLines('Genes that are differentially expressed at a FDR <= 0.05 in 6/6 RNASeq datasets')
paste(rownames(Adj.P.Value)[which(rowSums(Adj.P.Value[,-(4:5)]<=0.05)==6)], collapse = ',')
```

### Minimum Pvalue meta analysis
```{r minP}
Pval = DEXP

Pmin = data.frame(hgnc_symbols = Pval[,1], Pmin = rowMin(data.matrix(Pval[,-(1)])))
Pmin$Pmin = p.adjust(Pmin$Pmin, method = 'fdr')

tmp = dplyr::filter(Pmin, Pmin <= 0.05) 
tmp = data.frame(No.of.Genes = dim(tmp)[1], 
                 GeneList = tmp %>%
                   dplyr::select(hgnc_symbols) %>%
                   unlist %>% paste(collapse = ','))
writeLines("Genes that are atleast differentially expressed at an FDR <= 0.05 in one study are")
kable(tmp)
```

```{r fisher.brown}
Pfb = data.frame(hgnc_symbols = Pval[,1], 
                 Pfb = p.adjust(fishersPvalue(Pval[,-(1)]), method = 'fdr'))

Pfb$Pfb = p.adjust(Pfb$Pfb, method = 'fdr')

tmp = dplyr::filter(Pfb, Pfb <= 0.05) 
tmp = data.frame(No.of.Genes = dim(tmp)[1], 
                 GeneList = tmp %>%
                   dplyr::select(hgnc_symbols) %>%
                   unlist %>% paste(collapse = ','))

writeLines("Genes that are differentially expressed in the meta analysis at an FDR <= 0.05 are")
kable(tmp)
```

@@

# Store diffexp in synapse
write.table(diffexp, file = 'ALLDiffExp.tsv', sep = '\t', row.names=F, quote=F)
OBJ = File('ALLDiffExp.tsv', 
           name = 'Differential Expression (AD and Control)', 
           parentId = CODE$properties$id)
OBJ = synStore(OBJ,
               used = as.character(c(covariates.id)),
               executed = thisFile,
               activityName = 'MetaAnalysis of AMP-AD reprocessed data')
```

>>>>

library(Vennerable)
library(colorfulVennPlot)

@@

library(doParallel)
library(foreach)

clust = makeCluster(14)
registerDoParallel(clust)

@@

### Get covariates/sample metadata from synapse
```{r covariates}
# covariates.id = c('ROSMAP' = 'syn6114444',
#                   'MSBB' = 'syn6131421',
#                   'MAYO' = 'syn6128267')
# SVA Normalised
covariates.id = c('ROSMAP' = 'syn6129368',
                  'MSBB' = 'syn7059296',
                  'MAYO' = 'syn6128630')

@@

covariates = rbindlist(covariates, use.names = T, fill = T, idcol = 'Study')
```

### Get logCPM counts from synapse
```{r counts}
# logcpm.id = c('ROSMAP' = 'syn6114447' ,
#               'MSBB' = 'syn6131419',
#               'MAYO' = 'syn6128269')
# Get SVA adjusted data
logcpm.id = c('ROSMAP' = 'syn6129370',
              'MSBB' = 'syn7059294',
              'MAYO' = 'syn6128632')

@@

### Get residual expression from synapse
```{r counts1}
# Get SVA adjusted data
resid.expr.id = c('ROSMAP' = 'syn6129372',
                  'MSBB' = 'syn7059289',
                  'MAYO' = 'syn6128634')

resid.expr = lapply(resid.expr.id, downloadFile) %>%
  lapply(function(x){
    x = dplyr::select(x, -hgnc_symbol)
    }) %>%
  join_all(type = 'full') %>%
  dplyr::select(Gene.ID, ensembl_gene_id, one_of(covariates$ID)) 
resid.expr[is.na(resid.expr)] = dplyr::select(resid.expr, -Gene.ID, -ensembl_gene_id) %>%
  min(na.rm=T)
```

### Download differential expression data
```{r download.data}
# diffexp.id = c('ROSMAP' = 'syn6114453',
#                'MSBB' = 'syn6131423',
#                'MAYO' = 'syn6128273')
# Get SVA adjusted differential expression
diffexp.id = c('ROSMAP' = 'syn6129374',
               'MSBB' = 'syn7059298',
               'MAYO' = 'syn6128636')

@@

diffexp$ROSMAP = diffexp$ROSMAP %>%
  filter(Comparison == 'cogdx1-cogdx4') %>%
  dplyr::select(ensembl_gene_id, logFC, P.Value, adj.P.Val) %>%
  # dplyr::select(Gene.ID, logFC, P.Value, adj.P.Val) %>%
  dplyr::mutate(Tissue = 'DLPFC', Contrast = 'NCI-AD')

diffexp$MSBB = diffexp$MSBB %>%
  dplyr::filter(Name == 'ND-SD') %>%
  dplyr::select(ensembl_gene_id, logFC, P.Value, adj.P.Val, BrainRegion, Name) %>%
  dplyr::mutate(BrainRegion = factor(BrainRegion, labels = c('BM_10' = 'FP', 'BM_22' = 'STG', 
                                                             'BM_36' = 'PHG', 'BM_44' = 'IFG'))) %>%
  dplyr::rename(Tissue = BrainRegion, Contrast = Name)

diffexp$MAYO = diffexp$MAYO %>%
  dplyr::filter(Name == 'Control-AD') %>%
  tidyr::separate(Gene.ID, c('ensembl_gene_id', 'position'), sep = '\\.') %>%
  dplyr::select(ensembl_gene_id, logFC, P.Value, adj.P.Val, BrainRegion, Name) %>%
  dplyr::rename(Tissue = BrainRegion, Contrast = Name)

diffexp = rbindlist(diffexp, use.names = T, fill = T, idcol = 'Study')

@@

# Add hgnc_symbols to the resid.mat matrix
resid.expr = resid.expr %>%
  left_join(Ensemble2HGNC)

# Add hgnc_symbols to the diffexp matrix
diffexp = diffexp %>%
  left_join(Ensemble2HGNC)

diffexp$Direction = 'UP'
diffexp$Direction[diffexp$logFC < 0] = 'DOWN'

writeLines('Differentially expressed at an fdr of 0.05')
diffexp %>%
  filter(adj.P.Val <= 0.05) %>%
  group_by(Tissue, Direction) %>%
  summarise(n = length(unique(ensembl_gene_id))) %>%
  spread(Direction, n)

writeLines('Differentially expressed at an fdr of 0.05 and fold change of 1.2')
diffexp %>%
  filter(adj.P.Val <= 0.05, (logFC >= log2(1.2) | logFC <= -log2(1.2))) %>%
  group_by(Tissue, Direction) %>%
  summarise(n = length(unique(ensembl_gene_id))) %>%
  spread(Direction, n)
```

@@

### Meta Analysis
```{r metaanalysis}
P.Value = diffexp %>% 
  # filter(Tissue %in% c('DLPFC', 'PHG', 'TCX')) %>%
  filter(ensembl_gene_id != '') %>%
  dplyr::select(ensembl_gene_id, P.Value, Tissue, Contrast) %>%
  tidyr::unite(Tissue.Contrast, Tissue, Contrast, sep = '_') %>%
  tidyr::spread(Tissue.Contrast, P.Value)
P.Value[is.na(P.Value)] = 1

@@

# Rank prod based analysis
expr = resid.expr[!duplicated(resid.expr$ensembl_gene_id),-c(1:2)]
rownames(expr) = resid.expr$ensembl_gene_id[!duplicated(resid.expr$ensembl_gene_id)]

cl = covariates$Status
names(cl) = covariates$ID
cl[cl == 'NCI'] = 'Control'
cl[cl == 'ND'] = 'Control'
cl[cl == 'SD'] = 'AD'
cl = cl[cl %in% c('Control','AD')] %>% droplevels 
cl.tmp = -as.numeric(cl)+2
names(cl.tmp) = names(cl)

origin = covariates$BrainRegion %>% as.numeric
names(origin) = covariates$ID
origin = origin[names(cl)]

expr = expr[, names(cl)]

rankprod.pfp = RPadvance(expr, cl.tmp, origin, num.perm = 100, gene.names = rownames(expr), plot=T, rand = 123456, huge=T)
rankprod.pval = apply(rankprod.pfp$pfp, 1, min) %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  dplyr::rename(rankprod.pval = DF)

metaAnal = join_all(list(min.pval, max.pval, fb.pval, rankprod.pval), type = 'full')
metaAnal$weight = -log10(metaAnal[,-c(1,6)]) %>% rowSums %>% rank
metaAnal = metaAnal %>% arrange(desc(weight))

writeLines('Number of genes differentially expressed between AD and control at an FDR of 0.05 are')
tmp = metaAnal %>% 
  dplyr::select(-weight) %>%
  gather(method, pval, -ensembl_gene_id) %>% 
  group_by(method) %>% 
  filter(pval <= 0.05) %>%
  dplyr::select(ensembl_gene_id) %>%
  unique %>%
  dlply(.(method), .fun = function(x){unique(x$ensembl_gene_id)})
kable(sapply(tmp, length) %>% as.data.frame)

tmp1 = Venn(tmp)

plotVenn3d(tmp1@IndicatorWeight[-(1),'.Weight'], labels = names(tmp))
```
### Heatmap of expression
```{r heatmap}
# Get the overlapping gene sets from all three methods
tmp.expr = t(scale(t(expr)))
tmp.expr = scale(tmp.expr)
tmp.expr = tmp.expr[intersect(intersect(tmp$fb.pval, tmp$min.pval), tmp$rankprod.pval),]

tmp.covariates = dplyr::select(covariates, BrainRegion, Status)
tmp.covariates$Status[tmp.covariates$Status == 'NCI'] = 'Control'
tmp.covariates$Status[tmp.covariates$Status == 'ND'] = 'Control'
tmp.covariates$Status[tmp.covariates$Status == 'SD'] = 'AD'
tmp.covariates = tmp.covariates %>% droplevels
rownames(tmp.covariates) = covariates$ID

h1 = HeatmapAnnotation(tmp.covariates)
Heatmap(tmp.expr, top_annotation = h1, show_row_names = F, show_column_names = F,
        show_row_dend = F, show_column_dend = F)
```

### Pathway enrichment analysis
```{r pathway.enrich}
setsToTest = diffexp %>%
  filter(adj.P.Val <= 0.05) %>%
  dlply(.(Tissue, Direction), .fun = function(x){
    unique(x$ensembl_gene_id)
  })

tmp1 = rankprod.pfp$pfp[intersect(intersect(tmp$fb.pval, tmp$min.pval), tmp$rankprod.pval),] <= 0.05
setsToTest$META.UP = rownames(tmp1)[which(tmp1[,1])]
setsToTest$META.DOWN = rownames(tmp1)[which(tmp1[,2])]

# Convert ensembl id to hgnc_symbols
setsToTest = lapply(setsToTest, function(x, Ensemble2HGNC){
  Ensemble2HGNC$hgnc_symbol[Ensemble2HGNC$ensembl_gene_id %in% x] %>%
    unique()
}, Ensemble2HGNC)

# Get Background genes
backgroundGenes = Ensemble2HGNC$hgnc_symbol[Ensemble2HGNC$ensembl_gene_id %in% rownames(expr)] %>% unique()

############################################################################################################
# Function to perform Fishers enrichment analysis
fisherEnrichment <- function(genesInSignificantSet, # A character vector of differentially expressed or some significant genes to test
                             genesInGeneSet, # A character vector of genes in gene set like GO annotations, pathways etc...
                             genesInBackground # Background genes that are 
){
  genesInSignificantSet = intersect(genesInSignificantSet, genesInBackground) # back ground filtering
  genesInNonSignificantSet = base::setdiff(genesInBackground, genesInSignificantSet)
  genesInGeneSet = intersect(genesInGeneSet, genesInBackground) # back ground filtering
  genesOutGeneSet = base::setdiff(genesInBackground,genesInGeneSet)

  pval = fisher.test(
    matrix(c(length(intersect(genesInGeneSet, genesInSignificantSet)),             
             length(intersect(genesInGeneSet, genesInNonSignificantSet)),
             length(intersect(genesOutGeneSet, genesInSignificantSet)),
             length(intersect(genesOutGeneSet, genesInNonSignificantSet))), 
           nrow=2, ncol=2), alternative="greater")
  OR = (length(intersect(genesInGeneSet, genesInSignificantSet)) * length(intersect(genesOutGeneSet, genesInNonSignificantSet))) / (length(intersect(genesInGeneSet, genesInNonSignificantSet)) * length(intersect(genesOutGeneSet, genesInSignificantSet))) %>% as.numeric()
  return(data.frame(pval = pval$p.value,
                    ngenes = length(genesInGeneSet),
                    noverlap = length(intersect(genesInGeneSet, genesInSignificantSet)),
                    Odds.Ratio = OR,
                    Genes = paste(intersect(genesInGeneSet, genesInSignificantSet), collapse = '|')))}
############################################################################################################

############################################################################################################
#### Get gene sets ####
# Download enrichr gene sets from synapse
GL_OBJ = synGet('syn4867851')
ALL_USED_IDs = c(ALL_USED_IDs, GL_OBJ$properties$id)
load(GL_OBJ@filePath)

gsets = c("BioCarta_2015", "GO_Biological_Process", "Reactome_2015", "TargetScan_microRNA", "WikiPathways_2015")
GeneSets.Enrichr = GeneSets[gsets]

# Download AD related gene sets from synapse
GL_OBJ = synGet('syn4893059');
ALL_USED_IDs = c(ALL_USED_IDs, GL_OBJ$properties$id)
load(GL_OBJ@filePath)

GeneSets.CM = list(CellTypeMarkers = GeneSets[grep('Zhang', names(GeneSets))],
                   AD = GeneSets['AD:GeneticLoci'])
GeneSets = c(GeneSets.Enrichr, GeneSets.CM)

# Filter gene list
GeneSets = filterGeneSets(GeneSets, backgroundGenes, minSize = 3, maxSize = 5000)

enrichResults = list()
for (name in names(setsToTest)){
  genesInModule = setsToTest[[name]]  
  enrichResults[[name]] = lapply(GeneSets,
                                    function(x, genesInModule, genesInBackground, fisherEnrichment, rownameToFirstColumn){
                                      tmp = as.data.frame(t(sapply(x, fisherEnrichment, genesInModule, genesInBackground)))
                                      tmp = rownameToFirstColumn(tmp,'GeneSetName')
                                      return(tmp)
                                    }, unique(genesInModule), unique(backgroundGenes), fisherEnrichment, rownameToFirstColumn) %>%
    rbindlist(use.names=TRUE, idcol = 'Category') %>%
    dplyr::mutate(fdr = p.adjust(pval, 'fdr'))
  writeLines(paste0('Completed ',name))  
}

# Write results to file
tmp1 = rbindlist(enrichResults, use.names = TRUE, idcol = 'ComparisonName', fill = TRUE) %>%
  filter(!is.na(Category))

tmp1$pval = unlist(tmp1$pval)
tmp1$ngenes = unlist(tmp1$ngenes)
tmp1$noverlap = unlist(tmp1$noverlap)
tmp1$Odds.Ratio = unlist(tmp1$Odds.Ratio)
tmp1$Genes = unlist(tmp1$Genes)

# Store logcpm in synapse
write.table(tmp1, file = 'enrich.tsv', sep = '\t', row.names=F, quote=F)
OBJ = File('enrich.tsv', 
           name = 'Pathway Enrichment Results (AD and Control)', 
           parentId = CODE$properties$id)
OBJ = synStore(OBJ,
               used = as.character(c(covariates.id)),
               executed = thisFile,
               activityName = 'MetaAnalysis of AMP-AD reprocessed data')
```

@@

# Store residual expression in synapse
write.table(expr, file = 'ALLExpr.tsv', sep = '\t', row.names=F, quote=F)
OBJ = File('ALLExpr.tsv', 
           name = 'Residual Expression (AD and Control)', 
           parentId = CODE$properties$id)
OBJ = synStore(OBJ,
               used = as.character(c(covariates.id)),
               executed = thisFile,
               activityName = 'MetaAnalysis of AMP-AD reprocessed data')

# Store diffexp in synapse
write.table(diffexp, file = 'ALLDiffExp.tsv', sep = '\t', row.names=F, quote=F)
OBJ = File('ALLDiffExp.tsv', 
           name = 'Differential Expression (AD and Control)', 
           parentId = CODE$properties$id)
OBJ = synStore(OBJ,
               used = as.character(c(covariates.id)),
               executed = thisFile,
               activityName = 'MetaAnalysis of AMP-AD reprocessed data')

# Store meta analysis in synapse
write.table(metaAnal, file = 'ALLMetaAnal.tsv', sep = '\t', row.names=F, quote=F)
OBJ = File('ALLMetaAnal.tsv', 
           name = 'Meta Analysis (AD and Control)', 
           parentId = CODE$properties$id)
OBJ = synStore(OBJ,
               used = as.character(c(covariates.id)),
               executed = thisFile,
               activityName = 'MetaAnalysis of AMP-AD reprocessed data')
```
